MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments \textbf{Supplementary Information}

### 2022-07-08 main/MSstatsPTM_Main.tex
updated spikein sections, methods, and results

\subsection*{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}

@@

\medskip \noindent {\bf Experimental Design:}  Luchetti et al. \cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter. Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S. flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. 11 samples were allocated to 1 TMT mixture in an unbalanced repeated measure design. All conditions had two biological replicates except for the Dox1hr condition, which was allocated one replicate.

@@

The basic two-sample $t$-test \todo{ref - ideally from this literature if you can find} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the protein.

@@

\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation}. Limma then fits a linear model using the transformed inputs.

\begin{equation}
\begin{aligned}
\hat{u}_{ij} = \beta_0 + c_{i}\beta_1 + + r_j\beta_2 + \epsilon^{\prime\prime}_{ij}, \text{where } \epsilon^{\prime\prime}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{\prime\prime 2}/w_j), i \in \{1, 2\}, j \in \{1, 2\}
\end{aligned}
\label{eq:limma_model}
\end{equation}
\todo{Essentially, Limma can fit all kinds of models - with or without the bio replicate, including the ones in the equations above, it depends on the user. The only difference from the approaches above is the EB moderation.  I am confused by $r_j$ do you mean $I_{biorep=j}$? Since replicate id takes continuous values, the model above is a regression with replicate id as a single predictor / 1 df - not sure that's right;  I am confused by $w_j$ - this is not how Limma model is specified. Also what are the $\beta$ and why we only have one of them? Let's check which versions of the models the references use and how Ting describes Limma in her paper. }

The goal of Limma is to increase the sensitivity of detection of differential abundance when the experiment has few biological replicates. This is done by means of Empirical Bayes moderation, i.e. by borrowing the information regarding the variation across all the modification sites. Specifically, \todo{OV: this should be another model equation, rather than the text}

For each modification site, the model yields the same estimate $\hat{\Delta}_{adj}$ as \eqref{ttest_model}, but a different standard error. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages empirical bayes moderation~\cite{Robbins1992} to adjust the model variance of the individual PTM using the shared variance of all PTMs modeled in the experiment. \todo{DK: I introduced a new subscript here for the PTM p. It feel awkward to add it across the other methods so I only included it in the moderation describtion.} For a single PTM $p$ with an estimated fold change $\hat{\Delta}_p$, the estimated variance is calculated as $var(\hat{u}^{\ast}_{pij}) = \sigma^2_p / w_{pj}$ where $\sigma^2_p$ is the variance resulting from the linear model of PTM $p$ and $w_{pj}$ is the global variance across all PTMs in replicate $j$. The statistical significance is then determined using $\hat{\Delta}_p$ and $\widehat{SE(\hat{\Delta}_p)}$ calculated using the adjusted variance $var(\hat{u}^{\ast}_{pij})$.

@@

%Limma uses linear models to test the null hypothesis that there is no difference in mean PTM abundances between conditions. It leverages Empirical Bayes moderation to share pooled variance information across individual modification models and moderate the individual residual variances. Using a linear model allows Limma to share variance information across conditions, providing a more accurate estimate~\cite{Ritchie_15a}. With respect to PTM analysis,  Limma fits a linear model using the adjusted summarized features, $d_{1+}^{\ast}$ and $d_{2+}^{\ast}$, to estimate , $\Delta$. This results in the same hypothesis as seen in Equation \ref{eq:ttest_null_hyp}. The variance for the estimation is moderated using information from the variance across all PTMs modeled. For the estimate of a single PTM $m$, $\Delta_m$, the variance is estimated as

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
\begin{eqnarray}
& y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
& \text{where } \sum\limits_{i=1}^{J} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\label{eq:msstats_model}
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as

@@

Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}. Using Empirical Bayes helps overcome cases where the number of biological replicates in each condition is small. Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using EB moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This is due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Since this simulation is a group comparisons design that doesn't require many variance components, it is favorable to Limma and t-test

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In noisy simulations, MSstatsPTM \todo{had a higher metric XXX than} the existing methods}

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. The proposed approach can handle low feature counts well due to using Tukey's Median Polish for summarization. The existing approaches summation approach require more observations to correctly summarize the intensities in the run, whereas TMP is more robust. \todo{more/better}

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change, as compared to $t$-test and $Limma$. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In the group comparison TMT experiment, MSstatsPTM corrected for bias \todo{OV: 'corrected for bias' is unclear. We talk about confounding. Maybe use a more specific term?}}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} The results of this experiment are summarized in~\figref{data4_plots}. \figref{data4_venn_diagram} shows the number of significant modified peptides before and after adjustment. More peptides lost than gained differential abundance after the adjustment. 3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. For the peptides that became insignificant in the adjusted model, their change in abundance was mainly due to changes in global protein abundance. In contrast, for peptides that became significant after adjustment, their true abundance change was masked by underlying changes in the unmodified protein. Both of these issues were corrected in adjustment, and the unconfounded abundance change is shown. An additional question that must be addressed is if the decrease in significant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became insignificant due to an increase in variance. Thus we can conclude that the drop off in significant peptides was due to changes in global protein abundance.

In~\figref{data4_profile_plot}, the modification of protein $GSDMD$ at site $K62$ showed the advantage of the proposed method. The modified peptide originally showed no abundance change between the infected 1 hour, 4 hour, and 6 hour conditions. This was contrasted with a strong negative change in the global profiling run between the same conditions. Looking at the Dox4hr vs Dox1hr conditions and modeling the modified peptide without adjusting for changes in the global profiling, the fold change was $-.501$ and the adjusted p-value was insignificant at $.0644$. After adjusting for changes in the global protein abundance, the fold change was much higher, $2.79$, and the adjusted p-value became significant, $5.25e-8$. In this case the effect of the modified peptide was strongly confounded with changes in the global protein. The proposed method allowed us to remove this confounding and estimate the true effect.

\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}

\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} The results of this experiment are summarized in Figure \ref{fig:data5_plots}. Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment. Again more PTMs became insignificant after adjustment than became significant. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method. Without adjustment, there was a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run showed a similar log fold change of $2.014$ between the same conditions. This indicated that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted p-value became insignificant, going from $.0009$ to $.248$. Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
%mix1-mix2  & 1                           & 1                        \\ \hline
%mix1-mix3  & 0                           & -1                         \\ \hline
%mix1-mix4  & 1                          & 0                         \\ \hline
%mix2-mix3  & -1                          & -2                        \\ \hline
%mix2-mix4  & 0                           & -1                        \\ \hline
%mix3-mix4  & 1                           & 1                         \\ \hline
%\end{tabular}
%\end{table}

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/benchmark_fig.png}
\caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}. (a) Relative amounts of human lysate, background {\it E. Coli} lysate, and spike-in peptides. Volumes of each mixture totaled 20mg, with 10mg of background {\it E. Coli} lysate used to normalize mix 3 and mix 4 to make all mixtures directly comparable. 2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4. (b) The expected fold change of the spike-in peptides in each comparison. ``Unadj. true FC'' is the fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue). This was calculated by directly comparing the amounts of spike-in in each condition. ``Adj. true FC'' is the fold change of the spike-in peptides when adjusting for changes in human lysate. This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.5\linewidth}
\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
\caption{}
\label{fig:data4_venn_diagram}
\end{subfigure}
\begin{subfigure}[c]{0.95\linewidth}
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{}
\label{fig:data4_profile_plot}
\end{subfigure}
\caption{Dataset 4: Human - Ubiquitination - 1mix-TMT. a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons. More PTMs became insignificant after adjustment than became significant. For the peptides that became insignificant in the adjusted model, their change in abundance was driven by changes in the global protein. In contrast, peptides that became significant after adjustment had their true abundance change masked by underlying changes in the unmodified protein. b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$. When looking at the summary of the modification and global protein it was clear the conditions follow different trends. Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot. This indicated the modification was confounded with changes in the unmodified protein.}
\label{fig:data4_plots}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
\includegraphics[height=.5\textwidth]{images/shig_venn_diagramm.png}
\caption{}
\label{fig:data5_venn_diagram}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{}
	\label{fig:data5_profile_plot}
	 \end{subfigure}
\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT. a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons. Again more PTMs became insignificant after adjustment then became significant. b) Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$. When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend. Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.}
\label{fig:data5_plots}
\end{figure}

>>>>

\subsection*{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}
\label{sec:exp_proc_dataset3}

@@

\medskip \noindent {\bf Experimental Design:}  Luchetti et al.~\cite{LUCHETTI2021} profiled human epithelial cells engineered to express IpaH7.8 under a dox inducible promoter. Uninfected cells were measured at 0 and 6 hours, while cells infected with {\it Shigella flexneri} ({\it S. flexneri}) bacteria were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. 11 samples were allocated to 1 TMT mixture in an unbalanced repeated measure design. All conditions had two biological replicates except for the Dox1hr condition, which was allocated one replicate.

@@

The basic two-sample $t$-test~\cite{kutner_etal_04a} is based on the model
\begin{eqnarray}
y^*_{ijk} = \hat{\mu}^{*}_{i} + \epsilon^{\ast}_{ij}, \ \epsilon^{\ast}_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^{*2}),\ i=1,\ldots,I,\ j=1,\ldots,J,\ k=1,\ldots,K
\label{eq:ttest_model}
\end{eqnarray}
The model allows us to estimate $\hat{\Delta}$ and its standard error.
Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution with $df=2J-2$ degrees of freedom in balanced designs.
Unfortunately, this approach is fundamentally flawed as it does not account for the confounding between changes in the PTM abundance and the overall changes in the abundance of the protein.

@@

\todo{DK: Current plan is to follow what Ting did for MSstatsTMT and keep this part brief but expand on EB methods in the supplementary.}\medskip \noindent The two versions of $t$-test \todo{OV: ANOVA?} above can be expanded with Empirical Bayes moderation in Limma~\cite{ Ritchie_15a, Schwammle2015, Smyth:2004, Smyth:2005, Zhu,Chappell:2021}. In particular, the ratio-based version which models the ratios $\hat{u}_{ij}$ in~\eqref{adj_estimation} can be used to answer the null hypothesis in \eqref{null_hyp}. Limma uses the transformed inputs to fit a linear model with a fixed condition and run effect for each PTM.

\begin{eqnarray}
& u_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ \sum\limits_{i=1}^{J} Run_j = 0, \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\label{eq:limma_model}
\end{eqnarray}

The method than uses Empirical Bayes moderation to combine the variance estimations across all PTMs modeled. This increases the degrees of freedom with which the individual PTM variances are estimated~\cite{Ritchie_15a}. The goal of Limma is to increase the sensitivity of detection of differential abundance when the experiment has few biological replicates.

@@

@@

$MSstats$~\cite{Choi:2014} and $MSstatsTMT$~\cite{Huang:2020} are a series of R/Bioconductor packages designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. Models in these packages takes as input a list of log$_2$-intensities $y_{ijk}$. For each protein, the feature log$_2$-intensities are first summarized into a single value per protein per run using Tukey's median polish~\cite{Tukey:1977}. These summarized values are then used as input to fit a flexible family of linear mixed-effects models \cite{McLean:1991, Faraway:2006, Bolker2009}. The specific model instances depend on the experimental design and data acquisition. For example, to model the unmodified protein in \figref{data-structure} $MSstats$ fits the following model
\begin{eqnarray}
& y_{ij} = \mu + Condition_i + Run_{j} + \epsilon_{ij}, & \\ 
& \text{where } \sum\limits_{i=1}^{I} Condition_i = 0,\ Run \mathop\sim^{iid} \mathcal{N}(0, \sigma^2_j), \epsilon_{ij} \mathop\sim^{iid} \mathcal{N}(0, \sigma^2) & \nonumber
\label{eq:msstats_model}
\end{eqnarray}
Parameters of the model are estimated using restricted maximum likelihood (REML)~\cite{Kenward}.
In turn, the parameters allow us to estimate the relevant quantity $\hat{\Delta}_{Protein}$ and its standard error. Where $\hat{\Delta}_{Protein}$ is estimated as

@@

Similarly to $Limma$, $MSstatsTMT$ includes an optional Empirical Bayes moderation of the standard error~\cite{Huang:2020}, which increases the sensitivity of the model when the number of biological replicates in each condition is small. Based on the model, the approach tests the null hypothesis in \eqref{conv_null_hyp}, by comparing the test statistic against the Student distribution.

@@

\figref{sim1_fdr} illustrated the high FDR observed on the simulated datasets when not accounting for the overall changes in protein abundance. After the adjustment, the FDR of all methods improved. In a clean simulation, recall and accuracy of $MSstatsPTM$ and $Limma$ performed similarly. As expected when not using Empricial Bayes moderation in MSstatsPTM, Limma performed slightly better in simulations with fewer biological replicates. In contrast, the performance of the two-sample $t$-test method lagged behind. This was due to the fact that the $t$-test only uses data within the groups of interest while ignoring the remainder of the data. In contrast, $MSstatsPTM$ and Limma leverage information across comparisons, resulting in improved power in simulations with more comparisons. Additionally, since this simulation is a group comparison design that doesn't require many variance components, it is favorable to Limma and t-test

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In noisy simulations, MSstatsPTM better calibrated the true fold change over the existing methods}

@@

In simulations with missing values and few features, $MSstatsPTM$ outperformed $t$-test and $Limma$. Changes in unmodified protein level still needed to be accounted for to control the FDR. Once controlling for changes in the unmodified protein, the proposed method outperformed the other methods, as seen in~\figref{sim2_acc}. The proposed method calibrated model accuracy and recall rate well, even when the number of replicates were low. The proposed approach can handle low feature counts due to using Tukey's Median Polish for summarization. The existing methods summation approach requires more observations to correctly summarize the intensities in the run, whereas TMP is robust to few measurements.

Additionally, when comparing the fold change estimation across all modified peptides, the proposed method showed a tighter distribution of estimated fold changes around the true fold change. Specifically, the inner quartile range (IQR) of the estimated fold change for the proposed method was on average 21.8\% smaller than $Limma$ IQR and 10.4\% smaller than $t$-test's IQR. This means that while the mean of the estimated fold changes was generally correct for all methods, the proposed approach correctly estimated the fold change more often across all PTMs. This fold change comparison can be seen in {\bf Supplementary Figure~\ref{supp-fig:fc_boxplot}}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In two group comparison TMT experiments, MSstatsPTM corrected for confounding with the unmodified protein}

The results of these experiments are summarized in~\figref{data4_plots} and~\figref{data5_plots}. \figref{data4_venn_diagram} and~\figref{data5_venn_diagram} show the number of differentially abundant modified peptides before and after adjustment. In both cases, more peptides lost than gained differential abundance after the adjustment. A question that must be addressed is if the decrease in differentially abundant peptides is due to the increased variance that comes from adjustment. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became statistically insignificant after adjustment. In other words, the fold change was the same between models but variance increased. When this test was applied on this experiment, only one peptide became statistically insignificant due to an increase in variance. Thus we can conclude that the drop off in differentially abundant peptides was due to changes in global protein abundance.

Interestingly, the global protein was more likely to drive changes in abundance rather than mask changes in the PTM, i.e. when the protein and PTM moved in opposite directions. This can be seen in \figref{data5_profile_plot}, where the global protein $TTP$ followed the same trend as the modification at $S178$. Originally the modificaiton was shown to be differentially abundant between KO\_Uninfect and KO\_Early, however this was due to confounding with the global protein. The proposed method corrects for these changes automatically, allowing us to see the true fold change, without having to do a side by side comparison.

Conversely, cases where the PTM and global protein move in opposite directions are interesting and should be highlighted, as the modification and protein had contradictory abundance changes. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the global $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks. Conversely, the ubiquitination of $GSDMD$ at site $K62$ was actually upregulated between the same conditions. This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, as can be seen in \figref{data4_profile_plot}, appearing to have little change between no dox and dox 4 and 6 hour. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox1hr and Dox4hr conditions. The proposed method allowed us to see modifications of interest, while filtering out those that would originally be false positives.

%\subsubsection*{In the two-mixture TMT experiment, MSstatsPTM removed confounding \todo{OV: between what and what? Is it similar to the previous section?}}

%\todo{This section just describes the figure. Could you emphasize why we see these results? What are the reasons for differences in performance?} The results of this experiment are summarized in Figure \ref{fig:data5_plots}. Figure \ref{fig:data5_venn_diagram} shows the number of significant modified peptides before and after adjustment. Again more PTMs became insignificant after adjustment than became significant.  Again we tested if the decrease in significant peptides was due to removing confounding with changes in the umodified protein or increased variance after adjustment was applied. This was tested by looking for modified peptides whose adjusted log fold change was within 10\% of the unadjusted log fold change but became insignificant after adjustment. When this test was applied, 548 peptides became insignificant due to an increase in variance. This is only 3.4\% of the total peptides that became insignificant after adjustment. Thus we can conclude that the drop off in significant peptides was mainly due to changes in global protein abundance.

%In Figure \ref{fig:data5_profile_plot} the profile plot of protein $TTP$ modified at site $S178$ showed the power of the proposed method. Without adjustment, there was a large positive log fold change of $2.9$ between the WT\_Late and WT\_Uninfected conditions. However, the global profiling run showed a similar log fold change of $2.014$ between the same conditions. This indicated that the abundance change in the modified peptide is nearly entirely due to changes in the global protein. When adjusting for the global protein the modified peptide's adjusted p-value became insignificant, going from $.0009$ to $.248$. Correcting for changes in global protein abundance allowed us to see the true impact of the modification at $S178$ which would have otherwise been challenging to perceive.

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true FC & Adj. true FC \\ %\hline \hline
%mix2 vs mix1  & -1 & -1 \\ \hline
%mix3 vs mix1  & 0 & 1 \\ \hline
%mix4 vs mix1  & -1 & 0 \\ \hline
%mix3 vs mix2  & 1 & 2 \\ \hline
%mix4 vs mix2  & 0 & 1 \\ \hline
%mix4 vs mix3  & -1 & -1 \\ \hline
%\end{tabular}
%\end{table}

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/benchmark_fig.png}
\caption{{\bf Dataset 3: Spike-in benchmark - Ubiquitination - Label-free}. (a) Relative amounts of human lysate, background {\it E. Coli} lysate, and spike-in peptides. Volumes of each mixture totaled 20mg, with 10mg of background {\it E. Coli} lysate used to normalize mix 3 and mix 4 to make all mixtures directly comparable. 2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4. (b) The expected log2 fold change of the spike-in peptides in each comparison. ``Unadj. true FC'' is the log2 fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue). This was calculated by directly comparing the amounts of spike-in in each condition. ``Adj. true FC'' is the log2 fold change of the spike-in peptides when adjusting for changes in human lysate. This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}

@@

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_fc.png}
\caption{}
\label{fig:spikein_boxplot}
\end{subfigure}
\begin{subfigure}[c]{0.825\linewidth}
\includegraphics[width=1\textwidth]{images/spike_in_volcano.png}
\caption{}
\label{fig:spikein_prop_volcano}
\end{subfigure}
\caption{Dataset 3: Spike-in benchmark - Ubiquitination - Label-free. a) Before adjustment the fold change of the spike-in peptides' were systematically different from the expected fold change in all models. After adjustment, this systemic difference was removed, however the inner quartile range of the Limma and $t$-test models was wider than the proposed method. b) Before adjustment the spike-in peptides (colored red) did not follow the expected log fold change; after adjustment, the spike-in peptides were more in line with expectation. Using Limma, the spike-in peptides followed the expected log fold change after adjustment, however the majority of spike-in peptides did not have a significant adjusted p-value.}
\label{fig:spikein_volcano}
\end{figure}

\begin{figure}[ht]
\centering
\begin{subfigure}[c]{0.5\linewidth}
\includegraphics[width=1\textwidth]{images/ipah_venn_diagramm.png}
\caption{}
\label{fig:data4_venn_diagram}
\end{subfigure}
\begin{subfigure}[c]{0.95\linewidth}
\includegraphics[width=1\textwidth]{images/IpaH_prof_plot.png}
\caption{}
\label{fig:data4_profile_plot}
\end{subfigure}
\caption{Dataset 4: Human - Ubiquitination - 1mix-TMT. a) The overlap of differential modified peptides for the PTM model with and without global protein level adjustment across all comparisons.  3,236 modified peptides became insignificant, 1,352 became significant, while 4,282 were significant in both models. b) Comparing the global profiling of protein $GSDMD$ with the ubiquitination of the protein at site $K62$. When looking at the summary of the modification and global protein it was clear the conditions follow different trends. Specifically, there appeared to be no change in abundance between Dox1hr and Dox4hr in the modified plot, however there was a large negative change when looking at the unmodified plot. This indicated the modification was confounded with changes in the unmodified protein.}
\label{fig:data4_plots}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}{\textwidth}
 \centering
\includegraphics[height=.5\textwidth]{images/shig_venn_diagramm.png}
\caption{}
\label{fig:data5_venn_diagram}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=1.0\textwidth]{images/No_Difference_Shigella_Profile_Plot}
	\caption{}
	\label{fig:data5_profile_plot}
	 \end{subfigure}
\caption{Dataset 5: Mouse - Phosphorylation - 2mix-TMT. a) The overlap of differentially modified peptides between the PTM model with and without global protein level adjustment across all comparisons. 19,286 peptides became insignificant, 4,947 became significant, and 41,552 were significant in both models. b) Comparing the global profiling of protein $TTP$ with the modification of the protein at site $S178$. When looking at the summary of the modification and global protein it was clear the difference between conditions followed the same trend. Specifically, there was a positive adjustment in abundance when comparing WT\_Uninfect to WT\_Late in both the modification and global profiling run. This indicated the movement was driven by changes in global protein that was only accounted for in the model after adjusting for global protein abundance change.}
\label{fig:data5_plots}
\end{figure}

### 2022-07-08 supplementary/ptm_sm.tex
updated spikein sections, methods, and results

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_msstatsptm_volcano}
\caption{The modeling results of the proposed method both before and after adjustment. The spike-in peptides are colored red and the background peptides are colored grey. All grey peptides are expected to be insignificant. Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) did not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike-in peptides were more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment these false positives were decreased considerably. \label{fig:spike_volcano_msstats}}
\end{figure}

@@

\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT}
\label{sec:ipah}

In this experiment Shigella ubiquitin ligase IpaH7.8 was shown to inhibit the protein gasdermin D (GSDMD) ~\cite{LUCHETTI2021}. Multiplex proteomics was used to quantify the abundance of total protein, and ubiquitination in human epithelial cells. Cells were either infected or uninfected with IpaH7.8-deficient Shigella flexneri and measurements were taken at different time periods. Uninfected cells were measured at 0 and 6 hours, while infected cells were measured at 1, 2, 4, and 6 hour increments, resulting in six total conditions. The experiment was unbalanced with two biological replicates per condition for all conditions except for infected 1 hour. The experiment was modeled as a group comparison, with different replicates measured at each time point. The experimental design can be seen in Table~\ref{table:ipah_design}.

>>>>

\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{sim_new/spike_in_msstatsptm_volcano}
\caption{The modeling results of the proposed method both before and after adjustment. The spike-in peptides are colored red and the background peptides are colored grey. All grey peptides are expected to be insignificant. Using the proposed method to model the benchmark experiment, the spike-in peptides (colored red) did not follow the expected log fold change before adjustment. After adjusting for changes in overall protein abundance the spike-in peptides were more in line with expectation. Additionally the background grey colored peptides showed many false positives before adjustment. After adjustment these false positives were decreased considerably. \label{fig:spike_volcano_msstats}}
\end{figure}

@@

\subsubsection{Dataset 4 : Human - Ubiquitination - 1mix-TMT~\cite{LUCHETTI2021}}
\label{sec:ipah}

The experimental design can be seen in Table~\ref{table:ipah_design}.
